Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics

### 2024-08-23 Thesis/04-chap4.qmd
mapped out the front parts
finish the chap 4, next is to polish it

Enhanced visual diagnostics is the second group, which offers advanced visual aids for interpreting diagnostic plots. A notable example is the `DHARMa` package [@dharma], which uses an innovative approach by fitting quantile regression on scaled residual plots. It compares the empirical 0.25, 0.5, and 0.75 quantiles in scaled residuals with their theoretical counterparts, making it particularly useful for highlighting deviations from model assumptions, detecting model violations such as heteroscedasticity and incorrect functional forms, and uncovering issues specific to generalized linear models and mixed-effect models, like over/under-dispersion. By offering these enhanced visualizations, `DHARMa` enables users to more easily identify potential issues in their regression models that might not be immediately apparent with standard diagnostic plots. In summary, this group of packages enhances the interpretability of diagnostic plots by drawing attention to critical elements such as trends, clusters, and outliers. They sometimes automatically perform conventional tests on these elements, displaying the results as annotations, labels, or text within the plot, thereby further reducing the likelihood of misinterpretation.

@@

The `feature_pca_plot()` method visualizes the results of the PCA. By default, it plots the first principal component on the x-axis and the second principal component on the y-axis, with points colored according to their origin, true residual plots, null residual plots, or bootstrapped residual plots. Users can customize the x and y axes by specifying symbols for the `x` and `y` arguments. Additionally, the `col_by_set` option can be disabled if you prefer not to use coloring.

@@

While the R package `autovi` aims to provide tools that can be extended to broader visual inference applications, `autovi.web` is only focus on to for providing  a straightforward and clean user interface. An overview of the graphical user interface of `autovi.web` is provided in @fig-autovi-web. This is the default view of the web application, and there are five regions that user can mainly interact with. Region 1 of @fig-autovi-web is a sidebar menu which can switch between the analysis page and the information page. The analysis page is the focus of this section.

Region 2 of @fig-autovi-web is a panel for data uploading and CSV type selection. Clicking the "upload CSV" button opens a window where the user can select a file from their local system. The data status displayed above the button provides information about the number of rows and columns in the current dataset. Additionally, there are two example datasets available beneath the "upload CSV" button: one is a lineup example using a CSV file with three columns, and the other is a single plot example using a CSV file with two columns. More details about these example datasets are be discussed in @sec-autovi-web-example.

@@

Region 8 of @fig-autovi-web-result2 displays the density plot for bootstrapped visual signal strength and null visual signal strength. The densities are shown in distinct colors that are friendly for colorblind users. A solid vertical line marks the visual signal strength of the true residual plot, while rug lines at the bottom of the plot provide a clearer view of individual cases. Below the plot, a text message indicates the number and percentage of bootstrapped residual plots that would be rejected by the visual test when compared to the null plots. Note that the bootstrapped residual plots in this application are generated differently from `autovi`. Since we do not have the R model object, we can not refit the regression model with bootstrapped data. Instead, we bootstrap the residuals of the true residual plot directly to obtain bootstrapped residual plots. As as result, this panel will disappear when the true residual plot is unknown.

@@

### Workflow

The workflow of `autovi.web` is designed to be straightforward, with numbered steps displayed in the main user interface as shown in @fig-autovi-web. Since the design details have already been covered in @sec-autovi-web-design, we will not repeat the functionality of each panel in this section.

For general users, the workflow starts with Step 1: uploading a CSV file by clicking the "upload CSV" button. A window will pop up, allowing the user to select the CSV file. The CSV file should contain at least two columns representing fitted values and residuals of a residual plot. If a lineup needs to be evaluated, the CSV file should contain an additional column for labels of residual plot.  In Step 2, the user specifies the CSV type based on the uploaded file, choosing between a single residual plot or a lineup of multiple residual plots. Step 3 involves selecting the appropriate columns for fitted values, residuals, and, optionally, labels of residual plots, from the dataset. In Step 4, the user sets the number of bootstrapped draws needed for the analysis; more draws provide a better estimate of the bootstrapped distribution but result in slower computation. The user should also set the seed to control the simulation. If the dataset contains a lineup and the true residual plot is known, select its label. Finally, clicking the play button will initiate the analysis, and the user can review the result panels.

### Example {#sec-autovi-web-example}

@@

>>>>

Enhanced visual diagnostics is the second group, which offers advanced visual aids for interpreting diagnostic plots. A notable example is the `DHARMa` package [@dharma], which uses an innovative approach by fitting quantile regression on scaled residual plots. It compares the empirical 0.25, 0.5, and 0.75 quantiles in scaled residuals with their theoretical counterparts, making it particularly useful for highlighting deviations from model assumptions, detecting model violations such as heteroscedasticity and incorrect functional forms, and uncovering issues specific to generalized linear models and mixed-effect models, like over/underdispersion. By offering these enhanced visualizations, `DHARMa` enables users to more easily identify potential issues in their regression models that might not be immediately apparent with standard diagnostic plots. In summary, this group of packages enhances the interpretability of diagnostic plots by drawing attention to critical elements such as trends, clusters, and outliers. They sometimes automatically perform conventional tests on these elements, displaying the results as annotations, labels, or text within the plot, thereby further reducing the likelihood of misinterpretation.

@@

The `feature_pca_plot()` method visualizes the results of the PCA. By default, it plots the first principal component on the x-axis and the second principal component on the y-axis, with points colored according to their origi, true residual plots, null residual plots, or bootstrapped residual plots. Users can customize the x and y axes by specifying symbols for the `x` and `y` arguments. Additionally, the `col_by_set` option can be disabled if you prefer not to use coloring.

@@

While the R package `autovi` aims to provide tools that can be extended to broader visual inference applicaitons, `autovi.web` is only focus on to for provdinign a straightforward and clean uesr interface. An overview of the grahpical user interface of `autovi.web` is provided in @fig-autovi-web. This is the default view of the web applicaiton, and there are five regions that user can mainly interact with. Region 1 of @fig-autovi-web is a sidebar menu which can switch between the anlysis page and the information page. The analysis page is the focus of this section.

Region 2 of @fig-autovi-web is a panel for data uploading and CSV type selection. Clicking the "upload CSV" button opens a window where the user can select a file from their local system. The data status displayed above the button provides information about the number of rows and columns in the current dataset. Additionally, there are two example datasets available beneath the "upload CSV" button: one is a lineup example using a CSV file with three columns, and the other is a single plot example using a CSV file with two columns. More details about these example datasets are be discussed in @sec-autovi-web-workflow.

@@

Region 8 of @fig-autovi-web-result2 displays the density plot for bootstrapped visual signal strength and null visual signal strength. The densities are shown in distinct colors that are friendly for colorblind users. A solid vertical line marks the visual signal strength of the true residual plot, while rug lines at the bottom of the plot provide a clearer view of individual cases. Below the plot, a text message indicates the number and percentage of bootstrapped residual plots that would be rejected by the visual test when compared to the null plots. Note that the bootstrapped residual plots in this application are generated differently from `autovi`. Since we do not have the R model object, we can not refit the regression model with bootstrapped data. Instead, we bootstrap the residuals of the true residual plot directly to obtain bootstrapped residual plots. As as result, this panel will disappear when the true residual plot is unknonw.

@@

### Workflow {#sec-autovi-web-workflow}

The workflow of `autovi.web` is designed to be straightforward, with numbered steps displayed in each panel. Since the design details have already been covered in @sec-autovi-web-design, we will not repeat the functionality of each panel here.

Two example datasets are provided by the web application, as mentioned in @sec-autovi-web-design. The single residual plot example uses the `dino` dataset from the R package `datasauRus` [@datasaurus]. The lineup example uses residuals from a simulated regression model that has a non-linearity issue. We will walk through the lineup example to further demonstrate the workflow of the web application.

As shown in @fig-autovi-web-workflow-example, to use the lineup example data, click the "Use Lineup Example" button. The data status will then update to show the number of rows and columns in the dataset, and the CSV type will automatically be selected to the correct option. Since the example dataset follows the variable naming conventions assumed by the web application, the columns for fitted values, residuals, and labels of residual plots are set automatically. If the user is working with a custom dataset, these options must be set accordingly. Regardless of the dataset, the user must manually select the label for the true residual plot, as the web application allows assessments without this label. The next step is to click the play button to start the assessment.

In the result panels shown in [@fig-autovi-web-workflow-example]D, the first row of the table is the most crucial to check, as it provides the visual signal strength and the rank of the true residual plot among the other plots. The summary text beneath the table provides the $p$-value, which can be used for quick decision-making. The lineup is for manual inspection, and the user should see if the true residual plot is visually distinguishable from the other plots, to confirm if the model violation is serious. The density plot in [@fig-autovi-web-workflow-example]E offers a more robust result, allowing the user to compare the distribution of bootstrapped visual signal strength with the distribution of null visual signal strength. Finally, the grayscale attention map can be used to check if the target visual features, like the non-linearity present in the lineup example, are captured by the computer vision model, ensuring the quality of the assessment.

```{r fig-autovi-web-workflow-example, out.width = "100%", fig.cap = 'The workflow for `autovi.web` using the lineup example dataset proceeds as follows: (A) To begin, the user clicks the "Use Lineup Example" button to load the example dataset, during which the data status and CSV type will be automatically updated. (B) The target columns are selected automatically, though the user must manually select the label for the true residual plot, as the web application permits assessment without this label. (C) The user initiates the assessment by clicking the run button. (D) The visual signal strength of the true residual plot is displayed in the first row of the table, with a summary text beneath the table providing the $p$-value to aid in decision-making. The lineup on the right allows for manual inspection. (E) The density plot helps verify if the bootstrapped distribution differs from the null distribution, while the attention map on the right offers insights into whether the computer vision model has captured the intended visual features of the true residual plot.'}
magick::image_read_pdf("figures/autovi_web_workflow.pdf")
```

@@

This chapter presents an advancement in regression diagnostics software through the development of the `autovi` package and its accompanying web interface, `autovi.web`, addressing a critical gap in the current landscape of statistical software. While regression tools are widely available, effective and efficient diagnostic methods have lagged behind, particularly in the field of residual plot interpretation.

The `autovi` R package, introduced in this chapter, automates the assessment of residual plots by incorporating a computer vision model, eliminating the need for time-consuming and potentially inconsistent human interpretation. This automation improves the efficiency of the diagnostic process and promotes consistency in model evaluation across different users and studies.

The development of the accompanying Shiny app, `autovi.web`, expands access to these advanced diagnostic tools. By providing a user-friendly interface, it makes sophisticated residual plot assessment accessible to a broader audience, including those who may not have extensive programming experience. This web-based solution effectively addresses the potential barriers to adoption, such as complex dependencies and installation requirements, that are often associated with advanced statistical software.

The combination of `autovi` and `autovi.web` offers a comprehensive solution to the challenges of residual plot interpretation in regression analysis. These tools have the potential to significantly improve the quality and consistency of model diagnostics across various fields, from academic research to industry applications. By automating a critical aspect of model evaluation, they allow researchers and analysts to focus more on interpreting results and refining models, rather than grappling with the intricacies of plot assessment.

The framework established by `autovi` and `autovi.web` opens up exciting possibilities for further research and development. Future work could explore the extension of these automated assessment techniques to other types of diagnostic plots and statistical models, potentially revolutionizing how we approach visual inference in statistical analysis more broadly.
